Evolutionary rescue by aneuploidy in tumors exposed to anti-cancer drugs

### 2025-01-16 ResponseLetter.tex

In case it helps, my understanding of multi-step rescue theory is that there was an early model for the emergence of epidemics via multiple mutations (Antia et al 2003 Nature, and some follow ups like Yates et al 2006 PRSB and Alexander $\&$ Day 2010 Interface), a very general treatment of emergence and rescue via multiple mutations (Iwasa et al 2004 JTB), a 2-locus 2-allele rescue model (Uecker $\&$ Hermisson 2016 Genetics), and multi-step rescue models on a fitness landscape (Martin et al 2013, Osmond et al 2020), all of which also use branching processes. Many of the results in this paper have been derived in the above studies. For example, eqn 3 is essentially the 1-step rescue probability from Orr $\&$ Unckless 2008 (Am Nat, also their eqn 3) and eqn 4 is essentially the emergence probability from Iwasa et al 2004 (their eqn 3) multiplied by the number of first step mutations that occur during the wildtypes decline. Even more similarly, Eqn 4 is a special case of the results in Osmond et al (2020), who also used the reasoning from fitness valley crossing (Weissman et al 2009) to derive approximations for the probability of 2-step rescue, with the added complication of a distribution of mutation effects. Setting the distribution of fitness effects to a dirac and assuming fast wildtype decline reduces the first two lines of their eqn 8 to the first two lines of eqn 4 (the final line of eqn 4 being the well known case of 1-step rescue). The de novo vs standing variance result (eqn 6) was pointed out in Orr $\&$ Unckless 2008 (right side of page 163) and given in Martin et al 2013 (eqn 3.8). The rescue and recurrence times (eqns 8 and 9) are perhaps the most novel results derived here, but have some close similarities with times derived in Orr $\&$ Unckless 2014 (see eqn 18 and text S2 for rescue time and eqn 23 for recurrence time).

\textcolor{blue}{
We owe a serious debt of gratitude to the reviewer for this comment. They have basically done all the work for us. We have incorporated a review of the literature in the Introduction (line XXX), and refer to these papers throughout the Results at the equations mentioned. The only slight changes we have made are omitting the follow-ups to Antia et al 2003 and instead including Iwasa et al 2003 Proc B, which we really should have mentioned before and is arguably the closest thing in the literature to this manuscript.
Because the results mentioned by the reviewer are scattered across multiple papers, and because they all follow from simple branching
process approximations, we re-derive all results here for our model.
}

@@

- results: I thought it was interesting that the second scenario in eqn 5 was the geometric mean of the other two scenarios. Is there some intution for why this is so?

\textcolor{blue}{We agree, but we couldn't come up with such intuition.}

- results: I thought it might be helpful to give a little more intution about the effect of simultaneously increasing the aneuploid birth and death rates (paragraph starting below eqn 5). The reason why this leads to a decrease in $T^*$ and pushes the system towards the second scenario is that the mutation rate has increased, meaning a lineage doesn't have to persist for as long to almost surely lead to a successful rescue mutant, while the expected persistence time hasn't changed, making it more likely that a rare long-lived aneuploid lineage will lead to rescue.

\textcolor{blue}{%change 51
We have added the recommended comment (line XXX): "Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to a decrease in $T^*$, pushing the system to the second scenario. This is because increasing $\lambda_a$ causes a decrease in $T^*$ because  it increases the effective mutation rate $v\lambda_a$ (as mutations mostly occur during division) and a lineage does not have to survive as long in order to generate a successful mutant."
}

### 2025-01-19 ResponseLetter.tex

- results: I thought it was interesting that the second scenario in eqn 5 was the geometric mean of the other two scenarios. Is there some intution for why this is so?

\textcolor{blue}{We agree, but we couldn't come up with such intuition.}

- results: I thought it might be helpful to give a little more intution about the effect of simultaneously increasing the aneuploid birth and death rates (paragraph starting below eqn 5). The reason why this leads to a decrease in $T^*$ and pushes the system towards the second scenario is that the mutation rate has increased, meaning a lineage doesn't have to persist for as long to almost surely lead to a successful rescue mutant, while the expected persistence time hasn't changed, making it more likely that a rare long-lived aneuploid lineage will lead to rescue.

\textcolor{blue}{%change 51
We have added the recommended comment (line XXX): "Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to a decrease in $T^*$, pushing the system to the second scenario. This is because increasing $\lambda_a$ causes a decrease in $T^*$ as  it increases the effective mutation rate $v\lambda_a$ (as mutations mostly occur during division) and a lineage does not have to survive as long in order to generate a successful mutant."
}

@@

- results: I don't get the logic at the end of this paragraph, that your results are consistent with aneuploidy conferring resistance by decreasing the division rate. My understanding is that your result says that, holding aneuploid growth rate constant, increasing aneuploid division rate helps until you reach scenario 2 (effectively stationary/critical aneuploids) but beyond that doesn't help. I don't see how reducing division rate helps rescue.

\textcolor{blue}{%change 52
We have removed the sentence about these results being consistent with experimental finding.
}

### 2025-01-19 ms.tex
edited breast cancer stuff; ms.bib

We will address these questions essentially by combining known results from the evolutionary rescue literature, and considering the quantities and parameter values that are relevant for cancer.
Because these results are scattered across multiple papers, and because they all follow from simple branching process approximations, we re-derive all results here for our model.
While our focus is on the application of evolutionary rescue theory to evolution of drug resistance in cancer, our model is a general one, and in the general evolutionary rescue context, this work can be seen as a characterization of the circumstances in which two-step processes contribute to rescue even when single-step rescue mutations are available.
This is similar to part of \citet{osmond_genetic_2020} but focused on a different fitness landscape.

@@

In some of the analysis, we consider breast cancer. We assume that the division and death rates of the aneuploid breast cancer cells in the presence of cisplatin is the same as those of the sensitive cells in the absence of the drug. 
The growth rate of breast cancer cells in the absence of drug is $0.0085$ per day \citep{spratt1996rates}. The doubling time of breast cancer cells in the absence of cellular death is 8.2 days \citep{rew2000cell}. Thus, the division rate of aneuploid cells is $\lambda_a=\log{(2)}/8.2=0.0845$ per day, and their death rate is $\mu_a=0.076$ per day.
For the division and death rates of sensitive cells, we turn to results described in \citet{salehi2021clonal}. We select TNBC-SA535 PDX clone H, which has a relative fitness of $\approx 1.011$. Thus, the selection coefficient is the ratio of the aneuploid and sensitive Wrightian fitness values minus 1 \citep{chevin2011measuring}. The Malthusian fitness is the log of the Wrightian fitness \citep{Wu2013a}. Thus, $\Delta_s = \log{0.9975} -0.0025$. 
We assume cisplatin affects cancer cells by increasing their death rate but does not affect the their division rate, hence $\lambda_s=0.08$ and $\mu_s=0.0825$ per day.
A similar analysis for TNBC-SA609 clone A and TNBC-SA1035 clone H from \citet{salehi2021clonal} gives $\lambda_s=0.08, \mu_s=0.12$ and $\lambda_s=0.08, \mu=0.1$, respectively.

@@

Otherwise, aneuploidy is frequent enough ($u\lambda_a > \max{\left(-\Delta_a, 1/T^*\right)}$) to affect the evolution of drug resistance. 
The threshold tumor size can be approximated by one of the following scenarios, depending on $\Delta_a T^*$, which represents the change in the aneuploid log-population size during the critical time,
\begin{equation}  \label{eq:N_a}
\begin{aligned}
N_a^* \approx 
  \frac{\abs{\Delta_s}}{u\lambda_s} \cdot \begin{cases}
    \frac{\abs{\Delta_a}}{v\lambda_a}  \frac{\lambda_m}{\Delta_m} ,&
  \Delta_a T^* \ll -1 \text{ (tolerant aneuploids)},\\ 
  %\left(\frac{\lambda_a}{v}  \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
  2\lambda_a T^* ,&
  -1 \ll \Delta_a T^* \ll 1  \text{ (stationary aneuploids)},\\ 
  \frac{\lambda_a}{\Delta_a} ,&
   \Delta_a T^* \gg 1 \text{ (resistant aneuploids)}.
  \end{cases}
\end{aligned}
\end{equation}
This equation is equivalent to eq.~A3 of \citet{iwasa2004evolutionary} and to eq.~8 of \citet{osmond_genetic_2020} with the distribution of fitness effects set to a delta function;
our ``tolerant'', ``stationary'', and ``resistant'' scenarios correspond to \citet{osmond_genetic_2020}'s ``sufficiently subcritical'', ``sufficiently critical'', and ``sufficiently supercritical'', respectively. %change 57
These approximations are accurate when compared to results of stochastic evolutionary simulations (\Cref{rescue_prob,rescue_threshold}). 
In the first scenario, the treatment effectively kills aneuploid cells but not as quickly as it kills sensitive cells. 
In the second scenario, aneuploid cells are sufficiently resistant, and the expected size of each aneuploid lineage is roughly 1.
In both of these scenarios, aneuploidy increases the probability of rescue by slowing or halting the decrease in the tumor population size, allowing more opportunities to produce resistant mutants. 
In the third scenario, aneuploid cells are sufficiently resistant for the population to re-grow the tumor without additional resistance mutations.
In our parameterization for melanoma and breast cancer, the value of $\Delta_a T^*$ is XXX. Therefore, in these specific cases, we are in the XXX scenario. %TODO fill XXX % change b

@@

Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to a decrease in $T^*$, pushing the system to the second scenario. This is because increasing $\lambda_a$ causes a decrease in $T^*$ because  it increases the effective mutation rate $v\lambda_a$ (as mutations mostly occur during division) and a lineage does not have to survive as long in order to generate a successful mutant. 
Overall, the threshold tumor size $N_a^*$ is unaffected by the division rate $\lambda_a$ (i.e., $d \lambda_a T^*/d\lambda_a = 0$). Thus, if aneuploid cells rapidly die due to the drug but compensate by rapidly dividing, increasing the division rate will \emph{not} facilitate adaptation. %change 51
This is consistent with experimental findings where aneuploidy confers resistance by decreasing the division rate which in turn decreases the death rate as certain drugs affect cancer cells during division~\citep{replogle2020aneuploidy}. %change 52

@@

As an example, we have parameterized our model using estimates for melanoma and triple-negative breast cancer cells (TNBC) under drug therapy from the literature. 
We find that in these cases we are in the XXX scenario. %TODO fill XXX % change B
We have also compared the probability of evolutionary rescue between melanoma and TNBC (black and color lines, respectively, in \Cref{rescue_prob}B).
This comparison suggests that TNBC has a higher probability of relapsing compared to melanoma, because smaller tumors have higher resuce probability. Indeed, the probability of relapse in TNBC patients is higher than 50\% in the first 3-5 years after diagnosis \citep{taushanova2023synchronous}, whereas in melanoma patients it is approximately 10-20\% in the first 5 years \citep{wan2022prediction,von_schuckmann2019risk}. % change A

@@

Comparing our results to the evolutionary rescue literature, we complement the studies by \citet{iwasa2003evolutionary} and \citet{osmond_genetic_2020} on multi-step rescue by deriving simple approximations for a different fitness landscape. 
Similar to \citet{osmond_genetic_2020}, we find that two-step rescues can be more likely than simple one-step rescues. 
The ``stationary'' scenario, in which the aneuploid growth rate is close to zero in the presence of drug, is particularly interesting.
Although this may seem like a small region of parameter space in a general model of evolutionary rescue (Fig.~\ref{rescue_threshold}A, inset), the example of certain aneuploidies in cancer suggests it could be biologically significant. % TODO if change B gives stationary scenario, mention it here; otherwise, say that other cancer might produce this

@@

\nolinenumbers
%\bibliographystyle{unsrtnat}
\bibliographystyle{agsm}
\bibliography{evo2022}

>>>>

We will address these questions essentially by combining known results from the evolutionary rescue literature, and considering the quantities and parameter values that are relevant for cancer.
Because these results are scattered across multiple papers, and because they all follow from simple branching process approximations, we re-derive all results here for our model.
While our focus is on the application of evolutionary rescue theory to the evolution of drug resistance in cancer, our model is a general one, and in the general evolutionary rescue context, this work can be seen as a characterization of the circumstances in which two-step processes contribute to rescue even when single-step rescue mutations are available.
This is similar to part of \citet{osmond_genetic_2020} but focused on a different fitness landscape.

@@

In some of the analysis, we consider breast cancer. 
\citet{salehi2021clonal} studied multiple TNBC PDX clones (triple-negative breast cancer patient-derived xenografts).
Their results suggest that aneuploidy and other copy number alterations provid resistance to the drug cisplatin.  
Thus, we assume that aneuploid breast cancer cells are are partially or even completely resistant, i.e., $\Delta_m \ge \Delta_a > 0$, and that the division and death rates of aneuploid cells in the presence of cisplatin is the same as those of  sensitive breast cancer cells in the absence of the drug. 
The growth rate of breast cancer cells in the absence of drug is $0.0085$ per day \citep{spratt1996rates}.
The doubling time of breast cancer cells in the absence of cellular death is 8.2 days \citep{rew2000cell}.
Thus, the division rate of aneuploid cells is $\lambda_a=\log{(2)}/8.2=0.0845$ per day, and their death rate is $\mu_a=0.076$ per day.
To determine the division and death rates of sensitive cells, we chose TNBC-SA535 clone H from \citet{salehi2021clonal}, which has a relative fitness of $\approx 1.011$. Thus, the selection coefficient is the ratio of the aneuploid and sensitive Wrightian fitness values minus 1 \citep{chevin2011measuring}. The Malthusian fitness is the log of the Wrightian fitness \citep{Wu2013a}. Thus, $\Delta_s = \log{0.9975} -0.0025$. 
We assume cisplatin affects cancer cells by increasing their death rate but does not affect the their division rate, hence $\lambda_s=0.08$ and $\mu_s=0.0825$ per day.
A similar analysis for TNBC-SA609 clone A and TNBC-SA1035 clone H from \citet{salehi2021clonal} gives $\lambda_s=0.08, \mu_s=0.12$ and $\lambda_s=0.08, \mu=0.1$, respectively. % TODO it does not make sense that when you change aneuploid clones what changes is the sensitive growth rate - the sensitive growth rate should be constant and the aneuploid clone growth rate should change...

@@

Otherwise, aneuploidy is frequent enough ($u\lambda_a > \max{\left(-\Delta_a, 1/T^*\right)}$) to affect the evolution of drug resistance. 
The threshold tumor size can be approximated by one of the following scenarios, depending on $\Delta_a T^*$, which represents the change in the aneuploid log-population size during the critical time,
\begin{equation}  \label{eq:N_a}
\begin{aligned}
N_a^* \approx 
  \frac{\abs{\Delta_s}}{u\lambda_s} \cdot \begin{cases}
    \frac{\abs{\Delta_a}}{v\lambda_a}  \frac{\lambda_m}{\Delta_m} ,&
  \Delta_a T^* \ll -1 \text{ (tolerant aneuploids)},\\ 
  %\left(\frac{\lambda_a}{v}  \frac{\lambda_m}{\Delta_m}\right)^{1/2} ,&
  2\lambda_a T^* ,&
  -1 \ll \Delta_a T^* \ll 1  \text{ (stationary aneuploids)},\\ 
  \frac{\lambda_a}{\Delta_a} ,&
   \Delta_a T^* \gg 1 \text{ (resistant aneuploids)}.
  \end{cases}
\end{aligned}
\end{equation}
This equation is equivalent to eq.~A3 of \citet{iwasa2004evolutionary} and to eq.~8 of \citet{osmond_genetic_2020} with the distribution of fitness effects set to a delta function;
our ``tolerant'', ``stationary'', and ``resistant'' scenarios correspond to \citet{osmond_genetic_2020}'s ``sufficiently subcritical'', ``sufficiently critical'', and ``sufficiently supercritical'', respectively. %change 57
These approximations are accurate when compared to results of stochastic evolutionary simulations (\Cref{rescue_prob,rescue_threshold}). 
In the first scenario, the treatment effectively kills aneuploid cells but not as quickly as it kills sensitive cells. 
In the second scenario, aneuploid cells are sufficiently resistant, and the expected size of each aneuploid lineage is roughly 1.
In both of these scenarios, aneuploidy increases the probability of rescue by slowing or halting the decrease in the tumor population size, allowing more opportunities to produce resistant mutants. 
In the third scenario, aneuploid cells are sufficiently resistant for the population to re-grow the tumor without additional resistance mutations.
In our parameterization for melanoma and breast cancer, the value of $\Delta_a T^*$ is over 500.
Therefore, in these cases, we are in the third scenario, in which aneuploidy provides resistance to the drug. % change b

@@

Interestingly, increasing both the aneuploid division rate, $\lambda_a$, and the aneuploid death rate, $\mu_a$, such that the growth rate $\Delta_a$ remains constant, leads to a decrease in $T^*$, pushing the system to the second scenario. This is because increasing $\lambda_a$ causes a decrease in $T^*$ as  it increases the effective mutation rate $v\lambda_a$ (as mutations mostly occur during division) and a lineage does not have to survive as long in order to generate a successful mutant. 
In the second scenario, the threshold tumor size $N_a^*$ is unaffected by the division rate $\lambda_a$ (i.e., $d \lambda_a T^*/d\lambda_a = 0$). Thus, if aneuploid cells rapidly die due to the drug but compensate by rapidly dividing, increasing the division rate will \emph{not} facilitate adaptation. %change 51
%This is consistent with experimental findings where aneuploidy confers resistance by decreasing the division rate which in turn decreases the death rate as certain drugs affect cancer cells during division~\citep{replogle2020aneuploidy}. %change 52

@@

As an example, we have parameterized our model using estimates for melanoma and triple-negative breast cancer (TNBC) cells under drug therapy from the literature. 
We find that in these cases we are in the third scenario, in which aneuploidy provides at least partial resistance to the drug. 
It remains to be seen which tumor type and drug combinations produce tolerant and stationary aneuploidies. % change B
We have also compared the probability of evolutionary rescue between melanoma and TNBC (black and color lines, respectively, in \Cref{rescue_prob}B).
This comparison suggests that TNBC has a higher probability of relapsing compared to melanoma, because smaller tumors have higher resuce probability. Indeed, the probability of relapse in TNBC patients is higher than 50\% in the first 3-5 years after diagnosis \citep{taushanova2023synchronous}, whereas in melanoma patients it is approximately 10-20\% in the first 5 years \citep{wan2022prediction,von_schuckmann2019risk}. % change A

@@

Comparing our results to the evolutionary rescue literature, we complement the studies by \citet{iwasa2003evolutionary} and \citet{osmond_genetic_2020} on multi-step rescue by deriving simple approximations for a different fitness landscape. 
Similar to \citet{osmond_genetic_2020}, we find that two-step rescues can be more likely than simple one-step rescues. 
The ``stationary'' scenario, in which the aneuploid growth rate is close to zero in the presence of drug, is particularly interesting.
Although this may seem like a small region of parameter space in a general model of evolutionary rescue (Fig.~\ref{rescue_threshold}A, inset), it could be biologically significant for some tumor under specific drug therapies.

@@

\nolinenumbers
%\bibliographystyle{unsrtnat}
\bibliographystyle{agsm}
\bibliography{ms}
